Analysing the Performance of Shallow-Water Models with the Roofline Model

### 2016-03-09 docs/papers/roofline_model/gocean_roofline.tex
More updates to roofline doc

We want to use the Roofline Model to establish the efficiency (or
otherwise) of two shallow-water benchmark codes that we have used in
the GOcean project. The first of these codes, Shallow, was originally
written by Paul Swarztrauber over 30 years ago and has been optimised
by various people over that time. In contrast, nemolite2d is a new,
un-optimised code that has been extracted from NEMO.

Although we have investigated how the performance of the \psykal
version of nemolite2d compares with that of the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\cite{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound.

@@

\begin{figure}
\includegraphics[width=120mm]{stream2_archer}
\caption{Results from running the STREAM2 benchmark on the E5-2697
  CPUs. Error bars indicate the maximum and minimum values obtained
  from 200 repeats of each case.  size. }
\label{FIG_stream2}
\end{figure}

@@

The likwid tool (\url{https://github.com/RRZE-HPC/likwid}) gives
access to HWPCs on Intel and AMD CPUs.
However, there are issues with the accuracy of these counters,
especially for FLOP-related measurements on Ivy Bridge architectures. A
tool is included in the likwid distribution that tests the accuracy of
HWPC-derived data by running micro-benchmarks for which the FLOP count
and memory traffic are known:
\url{https://code.google.com/p/likwid/wiki/IvyBridgeVerification}.
The results of doing this on an Intel Xeon E5-1620 v2 CPU are shown in
Figure~\ref{FIG_flops_dp_test}. It appears that once the memory
footprint spills out of L1 cache (which all of our nemolite2d
benchmark configurations do), the performance counter results
significantly overestimate the FLOPs performed.

@@

We can also count the number of floating-point arithmetic operations
(FLOPs) in this code fragment which gives us 12 multiplications and 12
additions. The $AI$ for this code fragment is then $24*7/640 = 0.26$
FLOPs/byte.  If we do the same analysis for the (longer) u-Momentum
kernel of nemolite2d then we find that there are 51 addition
operations, 56 multiplication/division operations and one call to {\tt
  sin()}. There are 31 distinct memory references. This gives an $AI$
of approximately $(107+13)*7/(31*64) = 0.42$ FLOPs/byte (where we
assume the call to {\tt sin()} costs 13 FLOPs). The results of this
analysis are summarised in Table~\ref{TAB_kernel_details}.

\begin{table}
\begin{tabular}{|l|c|c|c|}
\hline
                             & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
WSS of time-stepping (bytes) & $104n^2$ & \multicolumn{2}{c|}{$180n^2$} \\ 
\hline
Kernel                     & 1st loop nest & Continuity & u-Momentum \\
\hline                                
FLOPs                      & 24      &    14      &   107+13   \\
Distinct memory refs       & 10      &    12      &   31       \\
$AI$ (FLOPs/byte)          & 0.26    &  0.128     &   0.42     \\
WSS (bytes)                & $56n^2$ &  $72n^2$   &   $136n^2$ \\
\hline
\end{tabular}
\caption{Details of the kernels investigated with the roofline
  model. WSS is the Working Set Size and $n$ is the problem size --
  the length of one side of the square domain.}
\label{TAB_kernel_details}
\end{table}

@@

Using these figures we can finally construct the Roofline model for
the two CPUs and the various kernels. In
Figure~\ref{FIG_roofline_archer} we plot the model for the E5-2697 CPU
and show the performance of the 1st loop-nest of Shallow and the
u-Momentum kernel of nemolite2d. Two things are immediately clear from
this plot: first, the performance of the first loop nest of Shallow is
clearly memory-bandwidth limited and second, the u-Momentum kernel of
NEMOLite2D is no-where near approaching this limit.

@@

Two performance points are plotted for the Shallow kernel
corresponding to domains of $512^2$ and $1024^2$. Using the equation
for working-set size in Table~\ref{TAB_kernel_details}, these
correspond to 14.7 and 58.7~MB, respectively. Since the E5-2697 has
an L3 cache of 30~MB, the $512^2$ case should be contained in cache
while the $1024^2$ case will spill to main memory. This is borne out
by the plot; the $1024^2$ case is achieving a performance in-line
with saturating the bandwidth to main memory. The $512^2$ case is
approaching the L3 bandwidth limit. We investigate the reasons for
the fact that it does not achieve the L3 bandwidth limit below.

@@

The continuity kernel includes array accesses of the form {\tt
  a(i,j-1)}. If array {\tt a} is allocated with extent {\tt (Nx, Ny)}
then this specifies a location in memory {\tt Nx} words before that
occupied by {\tt a(i,j)}. In order for such an access to be aligned,
{\tt Nx} must be a multiple of the SIMD vector width (2 for SSE, 4 for
AVX) and this must be specified to the compiler. We therefore padded Nx
as required and added a directive within the wrapper routine:
\begin{verbatim}
!DIR$ ASSUME (MOD(Nx,ALIGNMENT) .EQ. 0)
\end{verbatim}
where {\tt ALIGNMENT} is an integer parameter declared to have the value two
when compiling with SSE and four with AVX.

@@

Once all of the above steps were complete, the only un-aligned array
accesses reported by the compiler were those of the form {\tt
  a(i-1,j)}. Without taking a physical copy of the array {\tt a}, such
unaligned accesses are unavoidable in a stencil-based
finite-difference code. However, in order to quantify how much these
accesses limit the performance of the kernel we experimented with
replacing e.g. {\tt a(i-1,j)} with {\tt a(i-ALIGNMENT,j)}. The
resulting kernel had no un-aligned accesses and performed
significantly better; actually reaching or exceeding the L3-bandwidth
roofline for domains $64^2$ -- $256^2$.

The final SIMD-related optimisations are concerned with what happens
at the beginning and end of a SIMD-vectorisable loop. If the loop
updates e.g. {\tt a(i,j)} and begins on a value of {\tt i} that
  results in unaligned memory accesses (e.g. {\tt i = 2}) then the
  compiler must insert a ``peel'' loop to handle those iterations
  required to reach a point where memory is accessed on a 64-byte
  boundary.  Similarly, if the trip-count of the resulting
  vectorisable loop is not an integer multiple of the SIMD vector
  width then the compiler must insert a ``remainder'' loop to handle
  those iterations left over.  We removed the remainder loop by
  ensuring that the length of the inner loop was a multiple of the
  vector width and adding an {\tt ASSUME} compiler directive
  (c.f. that for {\tt Nx} above). This was straightforward since we
  had already padded the array extents themselves and NEMOLite2D uses
  a mask to define the active computational domain - the additional
  grid points resulting from the padding were simply marked as
  external to the domain.  Despite doing this we found that the
  remainder loop was not removed until we manually introduced the peel
  loop (rather than relying on the compiler).

>>>>

We want to use the Roofline Model to establish the efficiency (or
otherwise) of two shallow-water benchmark codes that we have used in
the GOcean project. The first of these codes, Shallow, was originally
written by Paul Swarztrauber over 30 years ago and has been optimised
by various people over that time. In contrast, NEMOLite2D is a new,
un-optimised code that has been extracted from NEMO.

Although we have investigated how the performance of the \psykal
version of NEMOLite2D compares with that of the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\cite{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound.

@@

\begin{figure}
\includegraphics[width=120mm]{stream2_archer}
\caption{Results from running the STREAM2 benchmark on the E5-2697
  CPU.}
\label{FIG_stream2}
\end{figure}

@@

The likwid tool (\url{https://github.com/RRZE-HPC/likwid}) gives
access to HWPCs on Intel and AMD CPUs.
However, there are issues with the accuracy of these counters,
especially for FLOP-related measurements on Ivy Bridge architectures. A
tool is included in the likwid distribution that tests the accuracy of
HWPC-derived data by running micro-benchmarks for which the FLOP count
and memory traffic are known:
\url{https://code.google.com/p/likwid/wiki/IvyBridgeVerification}.
The results of doing this on an Intel Xeon E5-1620 v2 CPU are shown in
Figure~\ref{FIG_flops_dp_test}. It appears that once the memory
footprint spills out of L1 cache (which all of our NEMOLite2D
benchmark configurations do), the performance counter results
significantly overestimate the FLOPs performed.

@@

We can also count the number of floating-point arithmetic operations
(FLOPs) in this code fragment which gives us 12 multiplications and 12
additions. The $AI$ for this code fragment is then $24*7/640 = 0.26$
FLOPs/byte.  If we do the same analysis for the (longer) u-Momentum
kernel of NEMOLite2D then we find that there are 51 addition
operations, 56 multiplication/division operations and one call to {\tt
  sin()}. There are 31 distinct memory references. This gives an $AI$
of approximately $(107+13)*7/(31*64) = 0.42$ FLOPs/byte (where we
assume the call to {\tt sin()} costs 13 FLOPs). The results of this
analysis are summarised in Table~\ref{TAB_kernel_details}.

\begin{table}
\begin{tabular}{|l|c|c|c|}
\hline
                             & Shallow & \multicolumn{2}{c|}{NEMOLite2D} \\
WSS of time-stepping (bytes) & $104n^2$ & \multicolumn{2}{c|}{$180n^2$} \\ 
\hline
Kernel                     & 1st loop nest & Continuity & u-Momentum \\
\hline                                
FLOPs                      & 24      &    14      &   107+13   \\
Distinct memory refs       & 10      &    12      &   31       \\
$AI$ (FLOPs/byte)          & 0.26    &  0.128     &   0.42     \\
WSS (bytes)                & $56n^2$ &  $72n^2$   &   $136n^2$ \\
\hline
\end{tabular}
\caption{Details of the kernels investigated with the roofline
  model. WSS is the Working Set Size and $n$ is the problem size --
  the length of one side of the square domain.}
\label{TAB_kernel_details}
\end{table}

@@

Using these figures we can finally construct the Roofline model for
the two CPUs and the various kernels. In
Figure~\ref{FIG_roofline_archer} we plot the model for the E5-2697 CPU
and show the performance of the 1st loop-nest of Shallow and the
u-Momentum kernel of NEMOLite2D. Two things are immediately clear from
this plot: first, the performance of the first loop nest of Shallow is
clearly memory-bandwidth limited and second, the u-Momentum kernel of
NEMOLite2D is no-where near approaching this limit.

@@

Two performance points are plotted for the Shallow kernel
corresponding to domains of $512^2$ and $1024^2$. Using the equation
for working-set size in Table~\ref{TAB_kernel_details}, these
correspond to 14.7 and 58.7~MB, respectively. Since the E5-2697 has an
L3 cache of 30~MB, the $512^2$ case should be contained in cache while
the $1024^2$ case will spill to main memory. This is borne out by the
plot; the $1024^2$ case is achieving a performance roughly in-line
with saturating the bandwidth to main memory. The $512^2$ case is
approaching the L3 bandwidth limit. We investigate the reasons for the
fact that it does not achieve the L3 bandwidth limit below.

@@

The continuity kernel includes array accesses of the form {\tt
  a(i,j-1)}. If array {\tt a} is allocated with extent {\tt (Nx, Ny)}
then this specifies a location in memory {\tt Nx} words before that
occupied by {\tt a(i,j)}. In order for such an access to be aligned,
{\tt Nx} must be a multiple of the SIMD vector width (2 for SSE, 4 for
AVX) and this must be specified to the compiler. We therefore padded {\tt Nx}
as required and added a directive within the wrapper routine:
\begin{verbatim}
!DIR$ ASSUME (MOD(Nx,ALIGNMENT) .EQ. 0)
\end{verbatim}
where {\tt ALIGNMENT} is an integer parameter declared to have the value two
when compiling with SSE and four with AVX.

@@

Once all of the above steps were complete, the only un-aligned array
accesses reported by the compiler were those of the form {\tt
  a(i-1,j)}. Without taking a physical copy of the array {\tt a}, such
unaligned accesses are unavoidable in a stencil-based
finite-difference code. However, in order to quantify how much these
accesses limit the performance of the kernel we experimented with
replacing e.g. {\tt a(i-1,j)} with {\tt a(i-ALIGNMENT,j)}. The
resulting kernel had no un-aligned accesses and performed
significantly better; actually reaching or exceeding the L3-bandwidth
roofline for domains of $52^2$ -- $256^2$ (Figure~\ref{FIG_cont_vec} -
``align all'').

The final SIMD-related optimisations are concerned with what happens
at the beginning and end of a SIMD-vectorisable loop. If the loop
updates {\it e.g.}\ {\tt a(i,j)} and begins on a value of {\tt i} that
results in unaligned memory accesses ({\it e.g.}\ {\tt i = 2}) then
the compiler must insert a ``peel'' loop to handle those iterations
required to reach a point where memory is accessed on a 64-byte
boundary.  Similarly, if the trip-count of the resulting vectorisable
loop is not an integer multiple of the SIMD vector width then the
compiler must insert a ``remainder'' loop to handle those iterations
left over.

We removed the remainder loop by ensuring that the length of the inner
loop was a multiple of the vector width and adding an {\tt ASSUME}
compiler directive ({\it c.f.}\ that for {\tt Nx} above). This was
straightforward since we had already padded the array extents
themselves and NEMOLite2D uses a mask to define the active
computational domain - the additional grid points resulting from the
padding were simply marked as external to the domain.  Despite doing
this we found that the remainder loop was not removed until we
manually introduced the peel loop (rather than relying on the
compiler). Figure~\ref{FIG_cont_vec} shows that the resulting binary
was the most performant of all those compiled with SSE on all problem
sizes that are contained within L3 cache with the exception of $52^2$.
{\bf Why is $52^2$ an exception? Something to do with tuning it to fit in L2?}

Since the caches of both of the CPUs used are eight-way associative,
we also considered whether we might be experiencing ``cache
thrashing'' whereby different locations in memory are mapped to the
same location in memory.
